### 2020-10-15 latex/acrobot.tex
More writing for acrobot 4.1-4.3

While this is the most general representation of an acrobot, the Hamiltonian
dynamics become too unwieldy for analytically proving results in this chapter.
We will therefore simplify the dynamics by assuming the acrobot is comprised of
two massless rods of equal length \(l\), with equal point masses \(m\) at the
tips (as in Figure \ref{fig:simple-acrobot-model}).
We call this type of acrobot a \textbf{simple} acrobot. We will also ignore any
frictional forces at both the hips and shoulder. 
Finally, it is important to note that a real gymnast cannot swing their hips in
full circles; 
for this reason, we assume that \(q_a \in [-\bar{q}_a, \bar{q}_a]\) where 
\(\bar{q}_a \in ]-\pi, \pi[\). 
If the acrobot is a robot as opposed to the model of a gymnast, 
\(\bar{q}_a\) can be used as a control parameter when designing VNHCs.

@@

Defining \(M(q) := D_s(q)\) and \(V(q) := P_s(q)\), we find the conjugate of momenta 
is \(p = (p_u, p_a) = M(q)\dot{q}\).
The dynamics in \((q,p)\) coordinates are given by
\begin{align}\label{eqn:acrobot-hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2}p\tpose \Minv(q) p -
    mgl\left(2 c_u + c_{ua}\right)\\
     &\begin{cases}
        \dot{q} = \Minv(q) p \\
        \dot{p}_u = -mgl\left(2s_u + s_{ua}\right) \\
        \dot{p}_a =-\frac{1}{2}p\tpose \nabla_{q_a}\Minv(q) p
        - mgl s_{ua} + \tau
    \end{cases} \nonumber
\end{align}
where the inverse inertia matrix is
\begin{equation}\label{eqn:Minv}
    \Minv(q) = \frac{1}{ml^2\left(2-c_a^2\right)}
    \begin{bmatrix}
        1 &
        -\left(1+c_a\right) \\
        -\left(1+c_a\right) &
        3+2c_a
    \end{bmatrix}
\end{equation}

\section{Previous Constraint Approaches}
% TODO: Give Xingbo's intuition, abstract away to qa = sin(theta) constraint
% based on VLP but show it doesn't work well in our framework
% TODO: Describe Xingbo's results and observe that his results were similar to a VNHC
% except they used VHC tools. We will modify his approach to use VNHCs and prove
% rigorously that we can stabilize any energy level set on the acrobot.
As we saw at the beginning of this chapter, there are some existing approaches
to generating giant motion for the acrobot.
We will examine a few of these approaches and examine whether they are viable
candidates on which to base a VNHC, akin to how we designed the VLP constraint
in Chapter \ref{ch:vlp}.

An instinctive approach to controlling the acrobot is to model it as a
variable-length pendulum as in Figure \ref{fig:acrobot-ecm}, 
since the length of its equivalent center of mass (ECM) changes with the leg motion. 
For instance, \citet{swingup_giant_acrobot} design a desired trajectory for the
ECM, then determine which leg angles \(q_a(t)\) are required to generate that
trajectory.

@@

Unfortunately, the VLP is not a true representation of the acrobot.
The effective length of the ECM is 
\[
    l_e(q_a) := l\sqrt{\frac{5}{4} + cos(q_a)}
    ,
\]
and its effective angle is
\[
    q_e := \arctan_2(s_u + \frac{1}{2}s_{ua}, -c_u - \frac{1}{2}c_{ua})
    .
\]
There are two things of note here. 
First, Figure \ref{fig:acrobot-vlp-symmetry} shows there are two configurations
of the acrobot which give the same effective angle and length.
This means the acrobot and the VLP are not equivalent representations;
designing a VNHC that injects energy using the ECM may not produce human-like
leg motion on the acrobot.
Second, the torque input \(\tau\) applied to \(q_a\) affects both
\(l_e\) and \(q_e\). 
In the VLP model from Chapter \ref{ch:vlp}, the control input only
affects the dynamics of the length variable.
If we wanted to design a VNHC to provably inject energy into the acrobot, we
would need to fully redesign the VLP constraint with a model where the control
input has a direct effect on the angle.

@@

Since the VLP representation of the acrobot requires redesigning a VLP
controller, and the resulting motion may not even be realistic giant motion,
this model is clearly ineffective for our purposes.

We turn to the thesis of \citet{xingbo_thesis}, who designs a VHC to enforce a
so-called ``tap" motion, with the purpose of injecting energy into the acrobot. 
First, he defines a compensator variable \(s\) which tracks \(\dot{q}_u\), so
that he can use the theory of VHCs with the extended configuration 
\((q_u,q_a,s)\).
He then finds \(h_1, h_2 \in \R_{>0}\) to define the
normalized radius \(\rho\) and normalized angle \(\xi\) in the
\((q_u, s)\)-plane.
These normalized variables are given by
\begin{align*}
    \rho &:= \sqrt{h_1 q_u^2 + h_2 s^2}
    , \\
    \xi &:= \arctan_2(h_2 s, h_1 q_u)
    . 
\end{align*}
He then sets the VHC to be \(h(q) = q_a - f_\text{rad}(\rho)f_\text{ang}(\xi)\)
with the control parameters \(\bar{q}_u\) and \(\rho_0\), where
\begin{align*}
    f_\text{rad}(\rho) &:= \tanh^2(\rho/\rho_0)
    , \\
    f_\text{ang}(\xi) &:= 
    \begin{cases}
        0 & -\pi < \xi \leq 0 \\
        \bar{q}_u \exp\left(1 - \frac{1}{1-(\frac{4\xi}{\pi} - 1)^2}\right) 
          & 0 < \xi \leq \frac{\pi}{2} \\
        0 & \frac{\pi}{2} < \xi \leq \pi
        .
    \end{cases}
\end{align*}

While his VHC shows promising experimental results, \citeauthor{xingbo_thesis}
does not provide an analytical proof that the acrobot gains energy.
His lack of analysis is tied to the fact that the constrained
(Lagrangian) dynamics are incredibly complicated.
In fact, it is even challenging to show that the constraint is regular.
While we could very easily convert his VHC into a VNHC by replacing \(s\) with
\(p_u\), we would run into the same problem. 
Since we want our constraint to \textit{provably} inject energy, we choose to
forgo this type of constraint.

>>>>

While this is the most general representation of an acrobot, the Hamiltonian
dynamics become too unwieldy for analytically proving results in this chapter.
To make the analysis tractable, we begin by assuming the acrobot is comprised of
two massless rods of equal length \(l\), with equal point masses \(m\) at the
tips (as in Figure \ref{fig:simple-acrobot-model}).
We call this type of acrobot a \textbf{simple} acrobot. 
We will also ignore any frictional forces at both the hips and shoulder. 
Finally, it is important to note that a real gymnast cannot swing their legs in
full circles; 
for this reason, we assume that \(q_a \in [-\bar{q}_a, \bar{q}_a]\) where 
\(\bar{q}_a \in ]-\pi, \pi[\). 
If the acrobot is a robot as opposed to the model of a gymnast, 
\(\bar{q}_a\) can be used as a control parameter when designing VNHCs.

@@

Defining \(M(q) := D_s(q)\) and \(V(q) := P_s(q)\), we find the conjugate of momenta 
is \(p = (p_u, p_a) = M(q)\dot{q}\).
The dynamics in \((q,p)\) coordinates are given by
\begin{align}\label{eqn:acrobot-hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2}p\tpose \Minv(q) p -
    mgl\left(2 c_u + c_{ua}\right)
    , \\
     &\begin{cases}
        \dot{q} = \Minv(q) p \\
        \dot{p}_u = -mgl\left(2s_u + s_{ua}\right) \\
        \dot{p}_a =-\frac{1}{2}p\tpose \nabla_{q_a}\Minv(q) p
        - mgl s_{ua} + \tau,
    \end{cases} \nonumber
\end{align}
where the inverse inertia matrix is
\begin{equation}\label{eqn:Minv}
    \Minv(q) = \frac{1}{ml^2\left(2-c_a^2\right)}
    \begin{bmatrix}
        1 &
        -\left(1+c_a\right) \\
        -\left(1+c_a\right) &
        3+2c_a
    \end{bmatrix}
    .
\end{equation}
The control input is a force \(\tau \in \R\) affecting only the dynamics of
\(p_a\), representing a torque acting on the hip joint.
This means \((q,p)\) are simply actuated coordinates, allowing us once again to
apply the theory of VNHCs from Chapter \ref{ch:vnhcs}.

Let us define the VNHC \(h(q,p) = q_a - f(q_u,p_u)\) of order 1, where
\(f \in C^2\left(\mathbb{S}^1 \times \R; \R\right)\).
To enforce regularity of \(h(q,p)\), we note that 
\(\nabla_{q_u}\Minv(q) = \Zmat{2\times 2}\).
Examining Theorem \ref{thm:vnhc-regularity}, our VNHC will be regular
if the regularity matrix
\[
    dh_q \Minv(q) \begin{bmatrix}0 \\ 1 \end{bmatrix}
    ,
\]
is of full rank on the constraint manifold \(\Gamma\).
Since 
\[
    dh_q = \begin{bmatrix}
        -\partial_{q_u} f & 1
    \end{bmatrix}
    ,
\]
the regularity matrix evaluates to the scalar equation
\[
    \frac{(1+c_a)\partial_{q_u}f(q_u,p_u) + (3+2c_a)}{ml^2(2-c_a^2)}
\]
which is full rank if and only if the numerator does not
change sign.
A sufficient condition for this is when \(f\) is a function solely of \(p_u\),
so that \(\partial_{q_u}f = 0\).
If this is the case, the regularity matrix is always positive and the VNHC 
\(h(q,p)\) is regular.

The acrobot is a noticeably more complex system than the VLP, as
the dynamics are coupled because \(\Minv(q)\) is not diagonal.
Because of this, the constrained dynamics of an arbitrary VNHC may not be easy
to determine analytically, which will impact our ability to prove energy
injection analytically.
In the rest of this chapter, our goal is to design the function \(f(q_u,p_u)\)
based on the natural human motion of a gymnast, with the caveat that the
resulting VNHC be regular and that we can prove the constrained dynamics will
inject energy into the acrobot.

\section{Previous Constraint Approaches}
% TODO: Give Xingbo's intuition, abstract away to qa = sin(theta) constraint
% based on VLP but show it doesn't work well in our framework
% TODO: Describe Xingbo's results and observe that his results were similar to a VNHC
% except they used VHC tools. We will modify his approach to use VNHCs and prove
% rigorously that we can stabilize any energy level set on the acrobot.
There are a few existing approaches to generating giant motion for the acrobot
which may be viable candidates on which to base a VNHC, akin to how we designed
the VLP constraint in Chapter \ref{ch:vlp}.

An initial approach to controlling the acrobot is to model it as a
variable-length pendulum by collapsing the two rods and masses into one
equivalent center of mass (ECM), as in Figure \ref{fig:acrobot-ecm}.
This seems a reasonable model reduction, since the length from the pivot to the
ECM changes depending on the angle \(q_a\) of the leg.
Indeed, \citet{swingup_giant_acrobot} use this approach to design a trajectory
for the ECM, then determine which leg angles \(q_a(t)\) are required to generate
that trajectory.

@@

Unfortunately, the VLP is not a true representation of the acrobot.
The effective length of the ECM is 
\[
    l_e(q_a) := l\sqrt{\frac{5}{4} + cos(q_a)}
    ,
\]
and its effective angle is
\[
    q_e := \arctan_2(s_u + \frac{1}{2}s_{ua}, -c_u - \frac{1}{2}c_{ua})
    .
\]
There are two important thing to note here. 
First, Figure \ref{fig:acrobot-vlp-symmetry} shows that for each pose of the
VLP representation, there are two configurations
of the acrobot which give the same effective angle and length.
This means the acrobot and the VLP are not equivalent representations;
designing a VNHC that injects energy using the ECM may not produce human-like
leg motion on the acrobot.

Second, if we were to compute the conjugate of momenta
\(p_{l_e}\) to \(l_e\) and \(p_e\) to \(q_e\), we would see the torque input
\(\tau\) appearing in both of their dynamic equations.
In the VLP model from Chapter \ref{ch:vlp}, the control input only
affects the dynamics of the length variable.
If we want to design a VNHC for this system, we cannot use any of the results
from Chapter \ref{ch:vlp} because the VLP models don't match.

@@

Since the VLP representation of the acrobot requires redesigning a VLP
VNHC, and the resulting ECM motion may not produce realistic leg motion,
this model is clearly ineffective for our purposes.

Let us turn to the thesis of \citet{xingbo_thesis}, who designs a VHC to enforce a
so-called ``tap" motion with the purpose of injecting energy into the acrobot. 
First, he defines a compensator variable \(s\) which tracks \(\dot{q}_u\), so
that he can use the theory of VHCs with the extended configuration 
\((q_u,q_a,s)\).
He then finds \(h_1, h_2 \in \R_{>0}\) to define the
normalized radius \(\rho\) and normalized angle \(\xi\) in the
\((q_u, s)\)-plane.
These normalized variables are given by
\begin{align*}
    \rho &:= \sqrt{h_1 q_u^2 + h_2 s^2}
    , \\
    \xi &:= \arctan_2(h_2 s, h_1 q_u)
    . 
\end{align*}
He then sets the VHC to be \(h(q) = q_a - f_\text{rad}(\rho)f_\text{ang}(\xi)\)
with the control parameters \(\bar{q}_u\) and \(\rho_0\), where
\begin{align*}
    f_\text{rad}(\rho) &:= \tanh^2(\rho/\rho_0)
    , \\
    f_\text{ang}(\xi) &:= 
    \begin{cases}
        0 & -\pi < \xi \leq 0 \\
        \bar{q}_u \exp\left(1 - \frac{1}{1-(\frac{4\xi}{\pi} - 1)^2}\right) 
          & 0 < \xi \leq \frac{\pi}{2} \\
        0 & \frac{\pi}{2} < \xi \leq \pi
        .
    \end{cases}
\end{align*}

While this constraint shows promising experimental results and it accurately
emulates true human motion, \citeauthor{xingbo_thesis}
does not provide an analytical proof that the acrobot will gain energy.
His lack of analysis is tied to the fact that the constrained
dynamics are incredibly complicated.
In fact, just showing the constraint is regular is a challenging task.
While we could very easily convert his VHC into a VNHC by replacing \(s\) with
\(p_u\), we would run into the same problem. 
Since we want our constraint to \textit{provably} inject energy, we must forgo
this type of constraint in favour of something less complex.
